Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}

### 2020-12-01 article_rvTH15_6.tex
up

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

@@

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.

Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.  
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat}.

@@

\caption{\label{fig:figs}concatenate of graphical units (GUs) is art}\label{fig:unnamed-chunk-1}
\end{figure}

But, as stated, in Prehistorical the writing system does not exists. Spatial relationships between graphical features, or graphical units, are not necessarly linear and directed but could most probably be more multi-directional and undirected: the direction of the interactions of pairwise GUs can be in any order.

@@

\caption{\label{fig:figs}Potential spatial relations between GUs.}\label{fig:unnamed-chunk-2}
\end{figure}

Because of the inherent variability of iconography, and because graphical and spatial proximities between GUs are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
  \item unexplicit groupings of graphical units -- like graphical units grouped into \emp{figures},  \emp{figures} grouped into \emp{patterns}, \emp{patterns} grouped into \emp{motives}, etc. -- with tedious number of groups
  \item consistency, proximities and relationships between these groups are often implicit and not quantified
  \item studies develop proper descriptive vocabularies, singular relationships of grouped, idosyncratic methods in a site-dependend or period-dependend scales 
\end{itemize}

These issues limit drastically the possibility to conduct cross-cultural comparisons and to draw a synthesis of humankind's symbolism at a large scale and over the long-term. 
\\
\\
In this article we present the \proglang{R} package \pkg{decorr}. Its purpose is to formalise a method based on relative neighborhood graphs to analyse any graphical content. As any formal system, iconography can be modelled as spatial features related one with the other depending on rules of spatial proximities, as state by the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70}. The principal idea of our model is that any graphical system can be represented by features connected (or not) to each other depending on their spatial proximity. To map and analyse these proximities, the model uses concepts and methods coming from Graph Theory. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) and their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by a relative neighborhood graphs analysis (Graph Theory) and a spatial (GIS) analysis.

Nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of graphical decorations, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. These \emph{x} and \emph{y} coordinates, measured in pixels, are relative to the decoration figure which is open in the first place in a new GIS project without any projection system. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis. The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The entire decoration image is tiled into different GUs' area of influence like a Voronoi diagram of the support where the Voronoi seeds are the GUs. Exist a link between a pair of GUs when their area of influence share a common border. A RNG -- also called a geometric graph, or a planar graph -- is constructed from GUs (nodes) and their proximity links (edges). The graph itself is as a subgraph of a Delaunay triangulation.

\begin{figure}[H]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age \emp{Cerro Muriano 1} stele (drawing: \cite{DiazGuardamino10}) with its extent (\code{xmin, xmax, ymin, ymax}); B) After the polygonisation of the GUs, including the border of the stelae, the Voronoi cells, the centroid of GUs and the links between GUs having adjacent cells (ie, sharing a border) are calculated; C) For each GUs, \emph{x} and \emph{y} are calculated; D) At least one variable, like the \code{type} of the GUs is defined in order to compute composition analysis. A simplier solution will be to create directly centroids (\code{POINTS}) on the GUs and to draw the edges manually}
\end{figure}

This model has a minimal of \emph{a priori} definitions. Those definitions only concern the intrinsic properties of GUs (type, technology, color, orientation, size, etc.) and the types of relations they share. Here, we will only consider one property for the nodes, its type (column \code{type}), and the most common types of relations between GUs: \emph{normal}, \emph{attribute} and \emph{overlapping} edges.

@@

It occurs frequently that a GU can be divided into a \emph{main unit} (eg, a character) and one or various \emph{attribute units} (eg, a helmet, male sex). Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of GUs. To record this information, a new type of edge, tagged with \code{'+'}, is be introduced. This type of edges is directed and, by convention, displayed with a dashed line. Its starts from the \emph{main unit} and ends with the \emph{attribute units}. The textual notation of such an edge is \code{'-+-'}. For example \code{4 -+- 6} means that the main node \code{4} has the attribute node \code{6}.

\begin{figure}[H] 
\centering
\includegraphics[scale=0.1]{gis_1}
\caption{\label{fig:gis1} GIS interface, Cerro Muriano 1 decoration. The GUs \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the GU \code{personnage} (anthropomorph).}
\end{figure}

@@

Finally, it is quite common that a graphical composition shows superimpositions between different UGs permit to distinguish different decoration phases for a single support. So, at first, the analyse must be performed on each different phases of decoration separately. This stratigraphical information (\code{A} \emph{over} \code{B}, or \code{B} \emph{under} \code{A}) helps to understand the relative chronology between GUs and must be recorded. A simple way to achieve this is to introduce the new tag \code{'>'} for the for the type of edge. This type of edges is directed. The textual notation of such an edge is \code{'->-'}. For example \code{A ->- B} means that \code{A} crosses \code{B} (ie, \code{A} overlaps \code{B} in the stratigraphical sense).

\begin{figure}[H] 
\centering
\includegraphics[scale=0.1]{gis_2}
\caption{\label{fig:gis1} GIS interface, Ibahernando decoration. The GUs \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

@@

The dataframe storing the inventory of decorations is \code{imgs}. The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})\\
%
% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Sun Nov 22 14:29:55 2020
\begin{table}[H]
\centering
\begin{tabular}{|p{0.5cm}|p{2.5cm}|p{2.5cm}|p{8.5cm}|}
  \hline
idf & site & decor & img \\ 
  \hline
  1 & Cerro Muriano & Cerro Muriano 1 & Cerro\_Muriano.Cerro\_Muriano\_1.jpg \\ 
    2 & Torrejon Rubio & Torrejon Rubio 1 & Torrejon\_Rubio.Torrejon\_Rubio\_1.jpg \\ 
    3 & Brozas & Brozas & Brozas.Brozas.jpg \\ 
    4 & Zarza de Montanchez & Zarza De Montanchez & Zarza\_de\_Montanchez.Zarza\_De\_Montanchez.jpg \\ 
    5 & Ibahernando & Ibahernando & Ibahernando.Ibahernando.jpg \\ 
   \hline
\end{tabular}
\caption{The studied corpus, the \code{imgs.tsv} dataframe} 
\label{Test_table}
\end{table}

%
The \pkg{decorr} package training dataset is composed by four stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package.
%
\begin{kframe}
\begin{verbatim}
R> # library("magick")
R> pth <- system.file("extdata", package = "decorr")
R> imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
+                     sep="\t", stringsAsFactors = FALSE)
R> lims <- list()
R> for(i in 1:nrow(imgs)){
+    i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
+    lbl.txt <- paste0(imgs[i,"idf"],"\n",
+                      imgs[i,"site"],"\n",
+                      imgs[i,"decor"],"\n",
+                      imgs[i,"img"],"\n",
+                      image_info(i1)$width,"*",image_info(i1)$height," px")
+    i1 <- image_annotate(i1,lbl.txt,location = "northwest",
+                         size = 25, color = "red")
+    lims[[length(lims)+1]]<- i1
+  }
R> out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]])),
+                            image_append(c(lims[[3]],lims[[4]]))),
+                          stack = TRUE)
R> plot(out.img)
\end{verbatim}
\end{kframe}\begin{figure}[H]

{\centering \includegraphics[width=\maxwidth]{figure/unnamed-chunk-5-1}

@@

\caption{\label{fig:figs}Decoration images of the training dataset}\label{fig:unnamed-chunk-5}
\end{figure}

@@

It contains the required minimum variables for the analysis.

\begin{kframe}
\begin{verbatim}
R> caption <- "Nodes (from \\code{nodes.csv} dataframe)}"
R> print(xtable::xtable(head(nodes),
+                       caption = caption),
+        table.placement="H")
\end{verbatim}
\end{kframe}% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Sun Nov 22 14:29:56 2020
\begin{table}[H]
\centering
\begin{tabular}{rl}
  \hline
 & V1 \\ 
  \hline
1 & site	decor	id	type	x	y \\ 
  2 & 1	Cerro Muriano	Cerro Muriano 1	1	personnage	349.814824824793	-298.324408661478 \\ 
  3 & 2	Cerro Muriano	Cerro Muriano 1	2	casque	349.814824824793	-243.98505280916 \\ 
  4 & 3	Cerro Muriano	Cerro Muriano 1	3	lance	238.463685783159	-298.324408661478 \\ 
  5 & 4	Cerro Muriano	Cerro Muriano 1	4	bouclier	446.022208956765	-381.169656108454 \\ 
  6 & 5	Cerro Muriano	Cerro Muriano 1	5	peigne	283.004141399813	-358.008619187794 \\ 
   \hline
\end{tabular}
\caption{Nodes (from \code{nodes.csv} dataframe)}} 
\end{table}

The primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The site is the current unit of analysis in Prehistory and Archaeology, but since a site can have various decorated objects, a primary key on two fields is necessary. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the GUs (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the correct the \emph{y} value of GUs nodes and edges, that is to say the \emph{y} value on the decoration image, the \pkg{decorr} calculate the absolute \emph{y} value and used the image height as a constant offset.

@@

The \code{edges} dataframe is quite similar to the \code{nodes} dataframe.

\begin{kframe}
\begin{verbatim}
R> caption <- "Edges (from \\code{edges.csv} dataframe)}"
R> print(xtable::xtable(head(edges),
+                 caption=caption,
+                 label="edgesdf"),
+        table.placement="H")
\end{verbatim}
\end{kframe}% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Sun Nov 22 14:29:56 2020
\begin{table}[H]
\centering
\begin{tabular}{rl}
  \hline
 & V1 \\ 
  \hline
1 & site	decor	a	b	type \\ 
  2 & 1	Cerro Muriano	Cerro Muriano 1	1	4	= \\ 
  3 & 2	Cerro Muriano	Cerro Muriano 1	1	5	= \\ 
  4 & 3	Cerro Muriano	Cerro Muriano 1	3	5	= \\ 
  5 & 4	Cerro Muriano	Cerro Muriano 1	1	2	+ \\ 
  6 & 5	Cerro Muriano	Cerro Muriano 1	1	7	+ \\ 
   \hline
\end{tabular}
\caption{Edges (from \code{edges.csv} dataframe)}} 
\label{edgesdf}
\end{table}

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common models in further studies, this direction helps to distinguish between nodes. The first column \code{edges$a} is the identifier of starting node or \emph{main node} or \emph{overlapping node}. The second column \code{edges$b} is the identifier of the ending node or \emph{attribue node} or \emph{overlapped node}. The \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the starting node and the ending node. There is no need to get the coordinates of the edges, these coordinates are calculated from the \code{nodes} dataframe. For example, Table \ref{edgesdf} shows that the first edge of the \emph{Cerro Muriano 1} decoration connects the nodes \code{1} and \code{8} (respectively in column \code{edges$a} and \code{edges$b}). A way to retrieve these connected nodes' coordinates will be:

\begin{kframe}
\begin{verbatim}
R> cm.1 <- subset(nodes, decor == "Cerro Muriano 1" & id == 1)[,c("x","y")]
\end{verbatim}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(e, x, parent.frame()): object 'decor' not found}}\begin{verbatim}
R> cm.8 <- subset(nodes, decor == "Cerro Muriano 1" & id == 8)[,c("x","y")]
\end{verbatim}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in eval(e, x, parent.frame()): object 'decor' not found}}\begin{verbatim}
R> cat(as.numeric(cm.1),";",as.numeric(cm.8))
\end{verbatim}

{\ttfamily\noindent\bfseries\color{errorcolor}{\#\# Error in cat(as.numeric(cm.1), "{};"{}, as.numeric(cm.8)): object 'cm.1' not found}}\end{kframe}

Once these three dataframes loaded, the list of decoration graphs can be calculated with the \code{list_dec()} function.

@@

The \code{list\_dec()} function allows to calculate graphs for all decorations stored into \code{nodes}, \code{edges} and \code{images}. The result is a list of decoration graph. The first graph of can be plotted

>>>>

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
% \setlist[itemize]{noitemsep, topsep=0pt}

@@

Symbolic practices is a characteristic trait of human societies. Even discussed, such practices seems to start between 233,000 to 800,00 BC \citep{dErrico00}, covering more than 97\% of total human societies time span. Symbolism cover a large range of practices, from ochre deposit in a tomb, to menhir alignements, passing through wall fresco. This latter, what might be called "iconographical practices", probably shows the most complex and interesting testimonies of past societies symbolism. For decades, its study was linked to history of religion because commonly seen as closely linked to cultual practices and believes. Since the \textit{New Archaeology} developpement during the 60's \citep{Clarke14}, symbolic expressions start to be studied with the same formal methods (statistics, seriations, distribution maps, etc.) as any another aspect of social organisation: settlement patterns, tools \emph{chaine op\'eratoire}, susbsitence strategies, etc. \citep{Renfrew91}. But unlike many aspects of the material culture where technological requirements and technical efficiency determine the choice of the raw material and the shape of the object -- a flint blade for cutting, a pottery for containing, a house for living --, the function of an iconographic composition cannot be drawn directly from itself. Whether these last decades study of ancient iconography had  undergone significative improvements at the site scale -- with GIS/database statistics, paleoenvironmental restitutions, etc. -- and at the sign scale with the development of archaeological sciences -- radiocarbon dating, use-wear analysis, elemental analysis, etc. --, these improvement do not necessarly help to understand the semantic content of the iconography.

Semantics or semiotics can be defined as a system of conventional features -- called signs -- organised also in conventional manners. Until our days, formal methods to study ancient iconography semantics have been mostlty been grounded  -- explicitly or not -- on the prime principle of Saussurian linguistic: the '\emph{linearity of the signifier}' \citep{Saussure89}.  
Writing is one of the most rational semiographical system with a clear distinction between signified and signifier and the development of the signified on a linear axis. Even if we do not understand the meaning of the signs, writing can easily be modelled with Graph theory and recurent patterns can be identified. For example, the 3-letters word "\code{art}" can be modelled with three vertices (\code{a}, \code{r}, \code{t}) and two edges (one between \code{a} and \code{r}, the other between \code{r} and \code{t}). In \proglang{R}, these features, concatenated in this order with a \code{paste0()}, is \code{art}, and not \code{rat}.
\smallbreak
\begin{figure}[H]

@@

\caption{\label{fig:figs}Concatenate of 'a', 'r' and 't' graphical units is 'art'}\label{fig:unnamed-chunk-1}
\end{figure}

But, as stated, in Prehistorical the writing system does not exists. Spatial relationships between graphical features, or graphical units, are not necessarly linear and directed but could most probably be more multi-directional and undirected: the direction of the interactions of pairwise graphical units can be in any order.

@@

\caption{\label{fig:figs}Potential spatial relations between graphical units}\label{fig:unnamed-chunk-2}
\end{figure}

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
  \item unexplicit groupings of graphical units -- like graphical units grouped into \emp{figures},  \emp{figures} grouped into \emp{patterns}, \emp{patterns} grouped into \emp{motives}, etc. -- with tedious number of groups
  \item consistency, proximities and relationships between these groups are often implicit and not quantified
  \item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependend or period-dependend scales 
\end{itemize}

These issues limit drastically the possibility to conduct cross-cultural comparisons and to draw a synthesis of humankind's symbolism at a large scale and over the long-term.  
\\
\\
In this article we present the \proglang{R} package \pkg{decorr}. Its purpose is to formalise a method based on relative neighborhood graphs (RNG) to analyse any graphical content (here: decorations). As any formal system, iconography can be modelled as spatial features related one with the other depending on rules of spatial proximities; as state by the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70}. And, because any graphical system is composed by features (mostly shapes) displayed spatialy, these systems can be represented by features connected (or not) to each other depending on their spatial proximity. To map and analyse these proximities, the model uses concepts and methods coming from Graph Theory, GIS and binary topological relationships (\emph{birel}) modeling. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) separated one with another by undecorated areas. Their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by a relative neighborhood graphs analysis (Graph Theory) and a spatial (GIS) analysis.\\
% \smallbreak
Nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').\\
The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So a decoration graph is a one component graph which is the dual graph of the Voronoi diagram (Fig. \ref{fig:gis}).

% \begin{figure}
\begin{figure}[H]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age \emp{Cerro Muriano 1} stelae (drawing: \cite{DiazGuardamino10}) opened in new GIS project without any projection system with its extent (\code{xmin, xmax, ymin, ymax} in pixels); B) Polygonsiation of the GUs, calculation of their centroids (nodes), calculation of the Voronoi cells, calculation of the proximity links (edges); C) plot of the relative \emph{x} and \emph{y} coordinates of the GUs centroids; D) plot of the values in the \code{type} of the GUs}
\end{figure}

Whether this theoritical process include image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.
\smallbreak

This model has a minimal of \emph{a priori} definitions. Those definitions only concern the intrinsic variables of GUs (\code{type}, \code{technologie}, \code{color}, \code{size}, etc.) and the types of relations they share. At least one variable, like the \code{type} of the GUs ('anthropomorph', 'horned helmet', etc.), is defined in order to compute composition analysis. Here, we will only consider one property for the nodes, its \code{type}, and the most common types of relations between GUs: \emph{normal}, \emph{attribute} and \emph{overlapping} edges.

@@

It occurs frequently that a GU can be divided into a \emph{main unit} (eg, a character) and one or various \emph{attribute units} (eg, a helmet, male sex). Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of GUs. To record this information, a new type of edge, tagged with \code{'+'}, is be introduced. This type of edges is directed and, by convention, displayed with a dashed line. Its starts from the \emph{main unit} and ends with the \emph{attribute unit}. The textual notation of such an edge is \code{'-+-'}. For example \code{4 -+- 6} means that the main node \code{4} has the attribute node \code{6} (Fig. \ref{fig:gis_1}).

\begin{figure}[H] 
\centering
\includegraphics[scale=0.05]{gis_1}
\caption{\label{fig:gis1} GIS interface, Cerro Muriano 1 decoration. The GUs \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the GU \code{personnage} (anthropomorph).}
\end{figure}

@@

Finally, it is quite common that a graphical composition shows superimpositions between different UGs permit to distinguish different decoration phases for a single support. So, at first, the analyse must be performed on each different phases of decoration separately. This stratigraphical information (\code{A} \emph{over} \code{B}, or \code{B} \emph{under} \code{A}) helps to understand the relative chronology between GUs and must be recorded. A simple way to achieve this is to introduce the new tag \code{'>'} for the for the type of edge. This type of edges is directed. The textual notation of such an edge is \code{'->-'}. For example \code{A ->- B} means that \code{A} crosses \code{B} (ie, \code{A} overlaps \code{B} in the stratigraphical sense) (Fig. \ref{fig:gis_2}).

\begin{figure}[H] 
\centering
\includegraphics[scale=0.05]{gis_2}
\caption{\label{fig:gis1} GIS interface, Ibahernando decoration. The GUs \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

These different kind of relationships are resumed with their birel equivalencies in the following table (Tab. \ref{table:tedges})

\begin{table}[H]
% \begin{center}
  \centering
 \begin{tabular}{|p{.5cm} p{.5cm} p{.5cm} p{2cm} p{2cm} p{5cm}|}
 \hline
  % node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
 GU.1 & edge type & GU.2 & (un)directed & \emph{birel} & description \\
 \hline
  \code{A} & \code{=} & \code{B} & undirected & $ A \cap B = \emptyset $ & \code{A} and \code{B} are disjoint \code{A} and \code{B} contemporaneous \\
  \code{A} & \code{+} & \code{B} & directed & $ A \cap B = A $ & \code{A} and \code{B} are contemporaneous, \code{B} is an attribute of \code{A} \\
  \code{A} & \code{>} & \code{B} & directed & $ A \cap B = \exists $ & \code{A} overlaps \code{B}, \code{A} is more recent than \code{B} \\
 % \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
 \hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs}\label{tab1}
% \end{center}
\label{table:tedges}
\end{table}

@@

The inventory of decorations is stored in the \code{imgs} dataframe. The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})\\
%
% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Thu Nov 26 18:04:41 2020
\begin{table}[H]
\centering
\begin{tabular}{|p{0.5cm}|p{2.5cm}|p{2.5cm}|p{8.5cm}|}
  \hline
idf & site & decor & img \\ 
  \hline
  1 & Cerro Muriano & Cerro Muriano 1 & Cerro\_Muriano.Cerro\_Muriano\_1.jpg \\ 
    2 & Torrejon Rubio & Torrejon Rubio 1 & Torrejon\_Rubio.Torrejon\_Rubio\_1.jpg \\ 
    3 & Brozas & Brozas & Brozas.Brozas.jpg \\ 
    4 & Zarza de Montanchez & Zarza De Montanchez & Zarza\_de\_Montanchez.Zarza\_De\_Montanchez.jpg \\ 
    5 & Ibahernando & Ibahernando & Ibahernando.Ibahernando.jpg \\ 
   \hline
\end{tabular}
\caption{The studied corpus, the \code{imgs.tsv} dataframe} 
\label{Test_table}
\end{table}

%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:timg})
%
\begin{kframe}
\begin{verbatim}
R> # library("magick")
R> pth <- system.file("extdata", package = "decorr")
R> imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
+                     sep="\t", stringsAsFactors = FALSE)
R> lims <- list()
R> for(i in 1:nrow(imgs)){
+    i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
+    lbl.txt <- paste0(imgs[i,"idf"],"\n",
+                      imgs[i,"site"],"\n",
+                      imgs[i,"decor"],"\n",
+                      imgs[i,"img"],"\n",
+                      image_info(i1)$width,"*",image_info(i1)$height," px")
+    i1 <- image_annotate(i1,lbl.txt,location = "northwest",
+                         size = 25, color = "red")
+    lims[[length(lims)+1]]<- i1
+  }
R> out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
+                            image_append(c(lims[[4]],lims[[5]]))),
+                          stack = TRUE)
R> plot(out.img)
\end{verbatim}
\end{kframe}\begin{figure}[H]

{\centering \includegraphics[width=\maxwidth]{figure/timg-1}

@@

\caption{\label{fig:figs}Decoration images of the training dataset}\label{fig:timg}
\end{figure}

@@

It contains the required minimum variables for the analysis (Fig. \ref{fig:tnds}).

\begin{kframe}
\begin{verbatim}
R> caption <- "Nodes (from \\code{nodes.csv} dataframe)}"
R> print(xtable::xtable(head(nodes),
+                       caption = caption),
+        table.placement="H")
\end{verbatim}
\end{kframe}% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Thu Nov 26 18:04:43 2020
\begin{table}[H]
\centering
\begin{tabular}{rllrlrr}
  \hline
 & site & decor & id & type & x & y \\ 
  \hline
1 & Cerro Muriano & Cerro Muriano 1 &   1 & personnage & 349.81 & -298.32 \\ 
  2 & Cerro Muriano & Cerro Muriano 1 &   2 & casque & 349.81 & -243.99 \\ 
  3 & Cerro Muriano & Cerro Muriano 1 &   3 & lance & 238.46 & -298.32 \\ 
  4 & Cerro Muriano & Cerro Muriano 1 &   4 & bouclier & 446.02 & -381.17 \\ 
  5 & Cerro Muriano & Cerro Muriano 1 &   5 & peigne & 283.00 & -358.01 \\ 
  6 & Cerro Muriano & Cerro Muriano 1 &   7 & sexe\_masculin & 342.69 & -427.49 \\ 
   \hline
\end{tabular}
\caption{Nodes (from \code{nodes.csv} dataframe)}} 
\end{table}

Since a prehistorical site can have various decorated objects, the primary key of the decoration is based on two fields: \code{nodes$site} and \code{nodes$decor}. The \code{nodes$id} is the identifier of the node. The \code{nodes$type} field is the default variable for further statistical analysis. Here, \code{nodes$type} refers to the typology of the GUs (anthropomorph, weapons, etc.). The \code{nodes$x} and \code{nodes$y} columns refer to the \emph{x} and \emph{}y coordinates of the nodes. As said, in the first place theses coordinates come from the GIS. But, in a GIS, the coordinates origin (0, 0) is the bottom-left corner and exist negative values, while this origin is top-left for any \proglang{R} matrices (rasters, grids, dataframes, etc.) with only positive values. To recover the local \emph{y} value of GUs nodes and edges on the decoration image, the absolute value \emph{y} value and the image height, as a constant offset, are computed.

@@

The \code{edges} dataframe is quite similar to the \code{nodes} dataframe (Fig. \ref{fig:teds}).

\begin{kframe}
\begin{verbatim}
R> caption <- "Edges (from \\code{edges.csv} dataframe)}"
R> print(xtable::xtable(head(edges),
+                 caption=caption,
+                 label="edgesdf"),
+        table.placement="H")
\end{verbatim}
\end{kframe}% latex table generated in R 3.6.1 by xtable 1.8-4 package
% Thu Nov 26 18:04:43 2020
\begin{table}[H]
\centering
\begin{tabular}{rllrrl}
  \hline
 & site & decor & a & b & type \\ 
  \hline
1 & Cerro Muriano & Cerro Muriano 1 &   1 &   4 & = \\ 
  2 & Cerro Muriano & Cerro Muriano 1 &   1 &   5 & = \\ 
  3 & Cerro Muriano & Cerro Muriano 1 &   3 &   5 & = \\ 
  4 & Cerro Muriano & Cerro Muriano 1 &   1 &   2 & + \\ 
  5 & Cerro Muriano & Cerro Muriano 1 &   1 &   7 & + \\ 
  6 & Cerro Muriano & Cerro Muriano 1 &   3 &   1 & = \\ 
   \hline
\end{tabular}
\caption{Edges (from \code{edges.csv} dataframe)}} 
\label{edgesdf}
\end{table}

Fields \code{edges$site} and \code{edges$decor} are the primary key of decoration. The fields \code{edges$a} and \code{edges$b} are the equivalent to columns \emph{from} and \emph{to} in Graph theory. Even if undirected graphs will the most common models in further studies, this direction helps to distinguish between nodes. The first column \code{edges$a} is the identifier of starting node or \emph{main node} or \emph{overlapping node}. The second column, \code{edges$b}, is the identifier of the ending node or \emph{attribue node} or \emph{overlapped node}. The \code{edges$type} is the type of relation (normal, attribute, overlapping, etc.) between the starting node and the ending node. There is no need to record the coordinates of the edges, these coordinates are calculated from the \code{nodes} dataframe. For example, Table \ref{teds} shows that the Cerro Muriano 1 decoration's first edge connects the nodes \code{1} and \code{8} (respectively in column \code{edges$a} and \code{edges$b}). A way to retrieve these connected nodes' coordinates will be:

\begin{kframe}
\begin{verbatim}
R> cm.1 <- subset(nodes, decor == "Cerro Muriano 1" & id == 1)[,c("x","y")]
R> cm.8 <- subset(nodes, decor == "Cerro Muriano 1" & id == 8)[,c("x","y")]
R> paste(as.numeric(cm.1),";",as.numeric(cm.8))
\end{verbatim}
\end{kframe}[1] "349.814824824793 ; 451.148926746119"  
[2] "-298.324408661478 ; -237.478157148203"

Once these dataframes loaded, the list of decoration graphs can be calculated with the \code{list_dec()} function.

@@

The \code{list\_dec()} function allows to calculate graphs for all decorations stored into \code{nodes}, \code{edges} and \code{images}. The result is a list of decoration graph. The first graph of the list can be plotted (Fig. \ref{fig:igraph})
